modelling

### 2019-03-29 old_book_files/_book/electoral-modelling-draft.tex
moved all old bookdown files to old_book_files

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-26}

@@

Consider the \texttt{Incomes} factor as an illustration. Independent of principal components, we may suspect that median personal income, median household income and median family income are providing similar information about the financial wellbeing of an electorate. Their loadings in the first principal component are large (0.20, 0.21 and 0.22 respectively), which provides the evidence needed to combine these variables into a single factor, which is called \texttt{Incomes}.

@@

\tiny 

@@

The spatial autoregressive coefficient \(\rho\) is positive and significant in only the 2001 and 2016 elections (Figure \ref{fig:rhovis}), meaning that in these elections, an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, it appears that after accounting for electoral socio-demographics, there is no significant spatial effect, meaning electorates effectively acted independently.

@@

{\centering \includegraphics[height=0.2\textheight]{electoral-modelling-draft_files/figure-latex/rhovis-1}

@@

@@

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect - the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time.

\emph{It is important here to note the ecological fallacy - insights are being drawn at the electorate level, and cannot be inferred for another disaggregate level (e.g.~individual voters).}

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Despite this, incomes have only significantly affected the electoral two-party preference in 2016 (see Figure \ref{fig:plotincomes}), with higher income areas associated with support for the Liberal party. In the prior elections, the estimated effect is still positive, albeit insignificant. Unemployment on the other hand, was influential in 2001, 2004 and 2007 elections (Figure \ref{fig:plotunemploy}), with electorates with higher unemployment favouring Labor. This estimated effect has become weaker over time, with unemployment having virtually no effect in 2016.

@@

{\centering \includegraphics[width=1\linewidth]{electoral-modelling-draft_files/figure-latex/plotincomes-1}

@@

\caption{Effect of Incomes}\label{fig:plotincomes}
\end{figure}

@@

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotunemploy-1}

@@

\caption{Effect of Unemployment}\label{fig:plotunemploy}
\end{figure}

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as \texttt{Extractive} industries) are consistently linked with higher support for the Liberal party, with the mangitude of this effect slightly increasing over the years (see Figure \ref{fig:plotextractive}). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (\texttt{Transformative}), as well as wholesale trade, retail trade, transport, post or warehousing (\texttt{Distributive}), are also linked with the Liberal party. However, these effects are not as strong, and \texttt{Transformative} was marginally insignificant in the 2010 election, whereas \texttt{Distributive} was not a key socio-demographic in 2001 or 2016. These effects are shown in Figures \ref{fig:plottransformative} and \ref{fig:plotdistributive}.

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotextractive-1}

\caption{Effect of Extractive}\label{fig:plotextractive}
\end{figure}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plottransformative-1}

\caption{Effect of Transformative}\label{fig:plottransformative}
\end{figure}
\begin{figure}[h]

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotdistributive-1}

\caption{Effect of Distributive}\label{fig:plotdistributive}
\end{figure}

Similarly, the proportions of workers in managerial, administrative, clerical and sales roles (\texttt{ManagerAdminClericalSales}) is also a significant predictor of two-party preference across all six elections, with higher proportions supporting the Liberal party.

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotmanager-1}

\caption{Effect of ManagerAdminClericalSales}\label{fig:plotmanager}
\end{figure}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotdiffaddress-1}

\caption{Effect of DiffAddress}\label{fig:plotdiffaddress}
\end{figure}

In each of the six elections, electorates with a higher proportion of people that have recently moved house (meaning in last five years) were more likely to support the Liberal party, with this effect becoming stronger since 2010. Having controlled for characteristics of house ownership and rental prices (via the factors \texttt{PropertyOwned} and \texttt{RentLoan} respectively), this is effect is somewhat surprising. \emph{Add more here?}

\hypertarget{indigenous-population}{%
\subsection{Indigenous population}\label{indigenous-population}}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotindigenous-1}

\caption{Effect of Indigenous population}\label{fig:plotindigenous}
\end{figure}

The indigenous population of an electorate is found to be a significant predictor in every election, with higher indigenous populations supporting the Liberal party. There are a few outliers with Lingiari (NT) being the most extreme. However, Lingiari itself has been a strong Labor supporter, and this effect appears to be driven by the other electorates with high Indigenous populations (Kalgoorlie (WA), Leichhardt (NSW), Durack (WA), Parkes(NSW) and Kennedy (QLD)), which have almost always elected a Liberal candidate.

\hypertarget{relationships-and-families}{%
\subsection{Relationships and families}\label{relationships-and-families}}

In each of the six elections, higher proportions people in de facto marriages are tied to support for the Labor party, whereas the effect of marriages themselves are insignificant. Larger family and household size (via the \texttt{FamHouseSize} factor) is also linked with Labor support, although only being significant in the 2007 and 2010 elections (see Figure \ref{fig:plotfamhousesize}).

@@

Many of the Census variables represent similar information, which is why factors were created and some variables were removed. However, there still remain some variables which are closely related. For example, \texttt{OtherLanguageHome} is the proportion of households that speak a language other than English at home, which is likely to be related to the proportion of people born overseas (a sum of the variables with a \texttt{Born} prefix). In 2001, the coefficient estiamte for \texttt{OtherLanguageHome} is significant and negative, and all \texttt{Born} variables are insignificant. If \texttt{OtherLanguageHome} is removed from the model, populations born in South Eastern Europe (\texttt{Born\_SE\_Europe}) absorb the negative effect (\(p = 0.0011\)).

@@

{\centering \includegraphics[width=0.75\linewidth]{electoral-modelling-draft_files/figure-latex/resstate-1}

@@

\caption{Boxplot of residuals by state.}\label{fig:resstate}
\end{figure}

\hypertarget{electorates-with-largest-residuals}{%
\subsection{Electorates with largest residuals}\label{electorates-with-largest-residuals}}

The electorates Capricornia (QLD), Leichhardt (QLD) and Riverina (NSW) have the three largest maximum absolute residuals across the six elections. Capricornia is a swing electorate, which strongly supported Labor between 2001 and 2010, before being narrowly won by Liberal in the last two elections. It has many people working in extractive jobs, an above-average Christian population, and above-average proportion of de facto relationships. The model tends to overpredict its two-party preferred vote (in favour of the Liberal party). Leichhardt is an area encompassing the northernmost tip of Queensland and consists of mainly small indigenous communities and uninhabited rural areas, which was won by the Liberal party in all but the 2007 election. It too has an above-average proportion of de facto relationships, along with an above-average proportion of people moving houses and a high proportion of single parent households. The model consistently underpredicts its two-party vote. Riverina is a rural electorate of NSW, and has been a very safe Liberal (National) seat since 1980. It has a relatively monocultural population consisting of a high proportion of Australian citizens and Christians, an above-average amount of extractive jobs and below average education levels. Its Liberal preference is also underpredicted.

\hypertarget{concluding-comments}{%
\chapter{Concluding comments}\label{concluding-comments}}

Across the past six elections, the spatial error models reveal that the electoral socio-demographics driving the two-party preferred vote have remained relatively steady, with industry and type of work being particularly influential, along with the type of relationships and household mobility, among others. The neighbourhood (spatial) effects were only found to be significant in the 2001 and 2016 elections. Surprisingly, unemployment, incomes and family/household sizes were significant in only some elections, and other characteristics like education were not significant at all.

The cross-sectional data set of Census and election data were obtained by collecting information from the ABS and AEC, and then applying a spatio-temporal imputation process, as detailed in section \ref{data}. This data is available in the \texttt{eechidna} \texttt{R} package, which can be downloaded from CRAN. All associated code to produce the data in the package from raw files can be found in the github repository (\url{https://github.com/ropenscilabs/eechidna}), and vignettes on how to use the data are located on the github page (\url{https://ropenscilabs.github.io/eechidna/}).

Furthermore, this paper is generated using the \texttt{knitr} \citep{knitr} package, so code to reproduce the results are contained in the \texttt{RMarkdown} file.

@@

\texttt{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) is an R package that makes it easy to look at the data from the Australian Federal elections and Censuses that occurred between 2001 and 2016. The data in this package includes voting results for each polling booth and electoral division (electorate), Census information for each electorate and maps of the electorates that were in place at the time of each event. All data in this package is obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government. Census data is imputed for the election years in which a Census does not occur via a spatio-temporal imputation method.

The Census and election datasets in the package can be linked by electorate using the unique identifier variable \texttt{UniqueID}. Vignettes describing how to use the data and how the data was obtained, can also be found on \url{https://ropenscilabs.github.io/eechidna/}.

>>>>

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-29}

@@

Consider the \texttt{Incomes} factor as an illustration. Independent of principal components, we may suspect that median personal income, median household income and median family income are providing similar information about the financial wellbeing of an electorate. Their loadings in the first principal component are large (0.19, 0.21 and 0.22 respectively), which provides the evidence needed to combine these variables into a single factor, which is called \texttt{Incomes}.

@@

\scriptsize 

@@

The spatial autoregressive coefficient \(\rho\) is positive and significant in only the 2001 and 2016 elections (Figure \ref{fig:rhovis}), meaning that in these elections, an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, the spatial effect weakens to become insignficant. In these years, it appears that the spatial component does not explain anything over and above the electoral socio-demographics, meaning electorates voted effectively independently.

@@

{\centering \includegraphics[height=0.2\textheight,]{electoral-modelling-draft_files/figure-latex/rhovis-1}

@@

\hypertarget{country-wide-trend}{%
\section{Country-wide trend}\label{country-wide-trend}}

Since all socio-demographics have been standardized to have a mean of zero and a variance of one, the intercept in each model can be interpreted as the estimated two-party preferred vote for an electorate with mean characteristics\footnote{Mean of all variables aside from Judaism, Indigenous, Islam and Buddhism, where it assumes the mean of the log value.}. The baseline of party preference has varied over the elections, with the biggest swing occuring in the 2007 election where the mean electorate shifted more than five percentage points in favour of the Labor party.

{\centering \includegraphics[height=0.2\textheight,]{electoral-modelling-draft_files/figure-latex/plotintercept-1}

\caption{Estimated intercept for each election, which represents the two-party preferred vote for an electorate with mean characteristics.}\label{fig:plotintercept}
\end{figure}

@@

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect - the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time in Figure \ref{fig:partresplot1} and Figure \ref{fig:partresplot2}.

It is important here to note the ecological fallacy - insights are being drawn at the electorate level, and cannot be inferred for another disaggregate level (e.g.~individual voters).

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps due to these policies, higher income electorates have been more likely to support the Liberal party, as the \texttt{Incomes} factor has a positive effect on Liberal preference (see row 1 in Figure \ref{fig:partresplot1}). This effect is signficant in every election aside from 2004, in which it is only marginally insignificant (\(p = 0.0613\)). Unemployment however, has not been as influential. In 2001 and 2004, electorates with higher unemployment were more likely to support Labor, but over time this has shifted towards support for the Liberal party, culminating in a significantly positive effect in 2016.

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as \texttt{Extractive} industries) are consistently linked with higher support for the Liberal party, with the mangitude of this effect slightly increasing over the years (see row 3 in Figure \ref{fig:partresplot1}). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (\texttt{Transformative}) are also more likely to support the Liberal party (see row 4 in Figure \ref{fig:partresplot1}).

Similarly, workers in managerial, administrative, clerical and sales roles (\texttt{ManagerAdminClericalSales}) is also a significant predictor of two-party preference vote across all six elections, with a higher proportion of people working these jobs increasing Liberal support. The magnitude of this effect has also increased over the years.

In each of the six elections, electorates with a higher proportion of people that have recently moved house (meaning in last five years) were more likely to support the Liberal party, although this effect was marginally insignificant in 2013 (see row 6 in Figure \ref{fig:partresplot1}. Having controlled for characteristics of house ownership and rental prices (via the factors \texttt{PropertyOwned} and \texttt{RentLoan} respectively), this is effect is somewhat surprising.

\hypertarget{relationships}{%
\subsection{Relationships}\label{relationships}}

De facto relationships, but not marriages, are found to be an important (and significant) predictor of the two-party preferred vote in all six elections, with more de facto relationships associated with higher support for the Labor party. Marriages however, are insignficiant.

\hypertarget{age}{%
\subsection{Age}\label{age}}

Older regions are often believed to be more conservative, and it can be seen that electorates with a higher median age have been more likely to support the Liberal party - although this effect is only significant in 2007 and 2010 (see row 2 in Figure \ref{fig:partresplot2}).

\hypertarget{education}{%
\subsection{Education}\label{education}}

Since 2007, electorates with higher education levels have been associated with supporting the Labor party, although this effect is only significant in 2016. Prior to 2007, education had an almost zero effect (see row 3 in \ref{fig:partresplot1}).

\hypertarget{diversity}{%
\subsection{Diversity}\label{diversity}}

Larger migrant populations from Asia, the Middle East, South-Eastern Europe, the United Kingdom and elsewhere, are either assocated with Labor support, or have no effect. Of these areas, only South-Eastern European populations were signficant in each election and Asian migrants were a significant in 2010. Speaking other language (aside from English) however, have a far stronger effect, as observed through \texttt{OtherLanguageHome}. Electorates with more diverse speech were more likely to support the Liberal party from 2004 onwards, with this effect being significant in 2007, 2010 and 2016. Furthermore, of the variables relating to religion, only Judaism shows a consistent effect, with electorates with larger Jewish populations more likely to vote Liberal.

\hypertarget{partial-residual-plots}{%
\subsection{Partial residual plots}\label{partial-residual-plots}}

@@

{\centering \includegraphics{electoral-modelling-draft_files/figure-latex/partresplot1-1}

@@

\caption{Partial residual plots for: income and unemployment, industry and type of work, and household mobility.}\label{fig:partresplot1}
\end{figure}

@@

{\centering \includegraphics{electoral-modelling-draft_files/figure-latex/partresplot2-1}

@@

\caption{Partial residual plots for: relationships and age, education and diversity.}\label{fig:partresplot2}
\end{figure}

@@

Many of the Census variables represent similar information, which is why factors were created and some variables were removed. However, there still remain some variables which are closely related. For example, electorate income levels (via \texttt{Incomes}) is likely to be related to electoral unemployment and labor force participation (via \texttt{Unemployment}). In 2001, the coefficient estiamte for \texttt{Unemployment} negative but not significant, and the \texttt{Incomes} variables is significant. If the \texttt{Incomes} variable is removed from the model, \texttt{Unemployment} absorbs the negative effect, becoming significant (\(p = 0.0056\)).

@@

{\centering \includegraphics[width=0.8\linewidth,]{electoral-modelling-draft_files/figure-latex/resstate-1}

@@

\caption{Boxplot of residuals by state with jittered points.}\label{fig:resstate}
\end{figure}

\hypertarget{outlier-electorates}{%
\subsection{Outlier electorates}\label{outlier-electorates}}

Based on the distribution of the cook's distance values, a cook's distance greater than \(0.1\) is considered to be influential and a potential outlier. The electorate of Sydney (NSW) has a large cook's distance from 2001 to 2013, due to its diverse population (language, birthplace and religion), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and Liberal vote is severely overpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in all but the 2007 election. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup - implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. Liberal support in Wentworth is underpredicted by the model in each year, and more so with Turnbull as Liberal leader.

Lingiari, an electorate taking up almost all of the Northern Territory, is an outlier in the 2001-2007 elections due to its large Indigenous population, young age profile and low rates of property ownership. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, and has very strong Labor support, making it influential in 2001, 2004 and 2010. Other electorates with large cook's distance are Barton (NSW) and Leichhardt (QLD) in 2016, and Canberra (ACT) in 2007.

\hypertarget{conclusion}{%
\chapter{Conclusion}\label{conclusion}}

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote have remained steady, whilst a few (typically weaker) effects have varied over time. Industry and type of work are particularly influential, with energy-related and manufacturing/construction jobs, as well as administrative roles being strongly linked with the Liberal party in all elections. Incomes have had a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment have shifted from weak association with Labor to a significant Liberal effect over the years, and higher education levels have been associated with Labor since 2007 (although only significant in 2016). It is also found that electorates with higher household mobility support Liberal, birthplace diversity favours Labor and more de facto relationships align with Labor preference - although marriages, family and household sizes have no material influence. Furthermore, the neighbourhood (spatial) effects were found to be positive in all elections, although only significant in 2001 and 2016, meaning that in the 2004-2013 elections, electorates effectively voted independently.

The findings in this paper complement the existing literature by modelling temporal trends, which as far as the authors are aware, has not been done previously for Australian elections using a regression framework. It is also the first study to model any Australian election since 2010 using Census information. Additionally, a key contribution of this research is the spatio-temporal imputation, which is applied to obtain Census data for the 2004, 2007, 2010 and 2013 elections. All data sets used in this study have been contributed to the \texttt{eechidna} \texttt{R} package, which provides a rich, accessible data resource for future Australian electoral analysis.

\hypertarget{acknowledgements}{%
\chapter{Acknowledgements}\label{acknowledgements}}

This paper was produced using \texttt{RMarkdown}\citep{rmarkdown} and \texttt{knitr} \citep{knitr}. All corresponding code for this paper can be found in the github repository (\url{https://github.com/jforbes14/journal-writing}), and the data used is available in the \texttt{eechidna} package \citep{eechidna}. All raw data was obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government.

@@

All election and Census data sets, along with electoral maps and more, are available in the \texttt{eechidna} \texttt{R} package, which can be downloaded from CRAN. \texttt{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) makes it easy to look at the data from the Australian Federal elections and Censuses that occurred between 2001 and 2016. This study contributed a large revision to the \texttt{eechidna} package, which included the addition of election and Census data for 2001-2010, vote outcomes for polling booths and imputed Census data for election years. For more details on using \texttt{eechidna}, please see the articles (vignettes) on the github page (\url{https://ropenscilabs.github.io/eechidna/}).

The authors would like to give a sincere thanks to Anthony Ebert, Heike Hofmann, Thomas Lumley, Ben Marwick, Carson Sievert, Mingzhu Sun, Dilini Talagala, Nicholas Tierney, Nathaniel Tomasetti, Earo Wang and Fang Zhou, all of whom have contributed to the \texttt{eechidna} package.

### 2019-03-29 old_book_files/raw-files/introduction.Rmd
moved all old bookdown files to old_book_files

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events, a Census occurs every five years, whereas an election usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling the raw data is outlined in @CHAPTER2, and the imputation process is detailed in @CHAPTER3. All data and associated documentation relating to this procedure has been made available in the `eechidna` `R` package on `CRAN`, providing a resource for future analysis.

Previous work on modelling voting outcomes with aggregate socio-demographics find that relatively good models can be fit. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 conduct principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. @Stimson09 model the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, but had little success. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that any temporal changes are observed, where each federal elections is modelled with a cross-sectional data set. Each cross-section consists of the two-party preferred vote (response variable), and socio-demographic variables (explanatory variables) that characterise each electorate. This process involves standardizing variables, using principal components to group variables into "factors" and finally fitting a spatial error model. These steps, and the rational behind a spatial model, are discussed in @CHAPTER4. Inference is then conducted on the models and insights are discussed in @CHAPTER5.

>>>>

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events, a Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling the raw data is outlined in @CHAPTER2, and the imputation process is detailed in @CHAPTER3. All data and associated documentation relating to this procedure has been made available in the `eechidna` `R` package on `CRAN`, providing a resource for future analysis.

Previous work on modelling Australian federal elections have found that aggregate socio-demographics are relatively good predictors of voting outcomes. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 use principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. On the contrary, @Stimson09 models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional data set. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from those previous.

The cross-sectional data set for each election consists of the two-party preferred vote (response variable), and socio-demographic variables (explanatory variables) that characterise each electorate. To obtain these cross-sections, socio-demographic variables are first standardized, and then principal components are used to group variables into "factors". To account for the inherent spatial structure of the data, a spatial error model is fit for each election. These steps are discussed in @CHAPTER4. In @CHAPTER5 inference and diagnostics are conducted on the models to see which effects are significant, how effects change over time and which electorates have abnormal voting behaviour.

### 2019-03-29 old_book_files/raw-files/modelling.Rmd
moved all old bookdown files to old_book_files

#### Dimension reduction
With only $N = 150$ observations (electorates) in each election and $p = 59$ socio-demographic variables in each cross-section, any model using all variables would face serious problems with multi-collinearity and over-fitting, likely leading to erroneous conclusions regarding variable significance. Therefore a form of dimension reduction is adopted that involves a two-stage approach that results in the identification of a reduced predictor set, for which models are fit.

@@

The spatial error model [@Anselin99] is chosen because it allows for spatial correlation in the errors. This can be thought of as capturing the political climate, where political pressures (unobserved characteristics influencing voter behaviour) of neighbours influence an electorate's voting behaviour. Spatial weights are calculated by allowing each electorate to be equally correlated with any electorate that shares any part of its boundary. The spatial error model is specified as follows:

Let $\rho$ be the parameter governing the extent of spatial correlation, $\boldsymbol v$ be a spherical error term, ${\bf W}$ be a matrix of spatial weights (containing information about the neighbouring regions), $\bf X$ be a matrix of socio-demographic covariates, $\boldsymbol \beta$ be a vector of regression coefficients and ${\boldsymbol u}$ be a vector of spatially correlated errors.

$${\bf y} = {\bf X} {\boldsymbol \beta} + {\boldsymbol u}$$

with

$${\boldsymbol u} = \rho {\boldsymbol W} {\boldsymbol u} + {\boldsymbol v}$$
where

@@

Estimation is done via maximum likelihood (using the `spdep` `R` package [@spdep]).

>>>>

#### Dimension reduction
With only $N = 150$ observations (electorates) in each election and $p = 59$ socio-demographic variables in each cross-section, any model using all variables would face serious problems with multi-collinearity and over-fitting, likely leading to erroneous conclusions regarding variable significance. Therefore a form of dimension reduction is adopted that involves a *two-stage approach* that results in the identification of a reduced predictor set, for which models are fit.

@@

The spatial error model [@Anselin99] is chosen because captures spatial heterogeneity by incorporating a spatially structured random effect vector [@LeSage2009]. In this context, the random effect thought of as capturing the political climate in each electorate, where the climate is correlated with the climate in neighbouring electorates. This functions under the assumption that the climate is independent of electoral socio-demographics, and that an electorate is equally correlated with any electorate that shares a part of its boundary. Spatial weights are calculated in accordance with these assumptions. The spatial error model is specified as follows:

Let $\rho$ be spatial autoregressive coefficient, $\boldsymbol v$ be a spherical error term, ${\bf W}$ be a matrix of spatial weights (containing information about the neighbouring regions), $\bf X$ be a matrix of socio-demographic covariates, $\boldsymbol \beta$ be a vector of regression coefficients and $\boldsymbol a$ be a spatially structured random effect vector.

$${\bf y} = {\bf X} {\boldsymbol \beta} + {\boldsymbol a}$$
and

$${\boldsymbol a} = \rho {\boldsymbol W} {\boldsymbol a} + {\boldsymbol v}$$
where

@@

Estimation is done using generalized least squares (using the `spdep` `R` package [@spdep]).

Table 4.1 details the resultant estimated model coefficients and their estimated standard errors for each of the six elections. These are interpreted in the next section.

*Model diagnostics are shown in Appendix \@ref(a:diagnostics), including the outcomes of tests for heteroskedasticity, residual non-normality, non-linear patterns in the residuals against included predictors, patterns in the residuals against missing predictors and influential points. These assessments did not indicate a major violation of any of the assumptions.*
